---
title: "Exercise set 5"
author: "Vira, Shulha"
date: "27/11/2022"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Exercise 1. Creating graphs and visualization

```{r E1-Q2.1}
## Add your code here
library(graph)
#library(help = "graph")
data("MAPKsig")
print(1)
numNodes(MAPKsig)
print(2)
numEdges(MAPKsig)
print(3)
edgeL(MAPKsig)$RALA
print(4)
adj(MAPKsig, index="BAD")
```



```{r hint-1}
## Creating a graphNEL that shows the distances among 3 major cities in Finland
V <- c("Tampere", "Helsinki", "Turku")
edL = list(Tampere=list(edges=c("Helsinki", "Turku"), weights=c(178, 164)), 
           Turku=list(edges=c("Tampere", "Helsinki"), weights=c(164, 169)),
           Helsinki=list(edges=c("Tampere", "Turku"), weights=c(178, 169))
           )
cities = graphNEL(nodes=V, edgeL=edL)
cities
plot(cities)
```



```{r hint-2}
## If value is equal to 7, we win, everything else we lose!
values = c(1:10)
result = list()
for (value in values){
  if (value == 7) {
    result[[value]] = "Win!"
  }
  else {
    result[[value]] = "Loss!"
  }
}
result
```



```{r E1-Q3.1}
## Add your code here
V = c("GO:0019395", as.list(GOBPCHILDREN["GO:0019395"])[[1]])
nnn = rbind(c(character(0)), c(character(0)), c(character(0)), c(character(0)),c(character(0)),c(character(0)),c(character(0)))
E = list()
for (i in 1:length(V)) E[[i]] = list(edges=c(character(0)))
names(E) = V
for (i in 1:length(V)){
  if (names(E)[i] != "GO:0019395"){
    E[[1]]$edges = append(E[[1]]$edges, names(E)[i])
  }
}
E
G = graphNEL(nodes=V, edgeL=E, edgemode='directed')
plot(G)
```



## Exercise 2. Pathway enrichment analysis (MAPK signaling pathway)

```{r E2-Q1.1}
## Add your code here
count_matrix = readRDS("/student_data/BBT.BI.202_2022/data/week_5/count_matrix_normalized_logged.RDS")
gene_list = readRDS("/student_data/BBT.BI.202_2022/data/week_5/DE_genes_PC_vs_BPH.RDS")
count_matrix
```



```{r E2-Q1.2}
## Add your code here
library(biomaRt)
mart = useDataset("hsapiens_gene_ensembl", useMart("ensembl"))

mappings_ensembl_entrez = getBM(
                  filters= "ensembl_gene_id",
                  attributes = c("entrezgene_id", "ensembl_gene_id"),
                  values=row.names(count_matrix), 
                  mart= mart)
mappings_ensembl_entrez
```



```{r E2-Q1.3}
library(org.Hs.eg.db)
map_ke <- as.list(org.Hs.egPATH)
kegg_entrez_pathway = map_ke[!is.na(match(mappings_ensembl_entrez$entrezgene_id, names(map_ke)))]

ensembl2kegg_pathway = list()
for (i in 1:nrow(mappings_ensembl_entrez)) {
  temp = kegg_entrez_pathway[as.character(mappings_ensembl_entrez[i, ]$entrezgene_id)]
  if (!is.na(names(temp))){
    ensembl2kegg_pathway[mappings_ensembl_entrez[i, ]$ensembl_gene_id] = temp
  } 
}
ensembl2kegg_pathway
```



### Pathway enrichment

```{r E2-Q1.4}
## Add your code here

```



#### Calculate N

```{r hint-3}
l = list(g1=c('p1'),
         g2=c('p2', 'p3'), # g2 is assigned to two ps.
         g3=c('p1'),
         g4= NA   # g4 is not assigned to any p.
         )
l
unlist(l) # check how the gene names are modified
sum(!is.na(unlist(l)))
```




```{r E2-Q1.6}
## Add your code here
N = sum(!is.na(unlist(ensembl2kegg_pathway)))
N
```



#### Calculate n for our pathway

```{r hint-4}
l = list(g1=c('p1'),
         g2=c('p2', 'p3'), # g2 is assigned to two ps.
         g3=c('p1'),
         g4= NA   # g4 is not assigned to any p.
         )

l_table = table(unlist(l))
l_table
l_table["p3"]
```



```{r E2-Q1.8}
## Add your code here
n = table(unlist(ensembl2kegg_pathway))["04010"][[1]]
n
```



#### Calculate M

```{r E2-Q1.10}
## Add your code here
ensembl2kegg_pathway_de = ensembl2kegg_pathway[names(ensembl2kegg_pathway) %in% rownames(gene_list)]
M = sum(!is.na(unlist(ensembl2kegg_pathway_de)))
M
```



#### Calculate k for our pathway

```{r E2-Q1.12}
## Add your code here
k = table(unlist(ensembl2kegg_pathway_de))["04010"][[1]]
k
```



#### Calculate enrichment P-values

```{r E2-Q1.14}
## Add your code here
contingency_table = matrix(nrow = 2, c(N - M - (n - k), M - k, n - k, k))
rownames(contingency_table) = c("Not significant", "Significant")
colnames(contingency_table) = c("Other GO terms", "GO term of interest")
contingency_table
a = fisher.test(contingency_table, alternative = "greater")
round(a$p.value, 2)
```



## Exercise 3. Pathway visualization

#### Calculate mean difference between PC and BPH used by pathview i.e. `gene.data`

```{r E3-Q2.1}
## Add your code here

count_matrix_mean_diff  = apply(count_matrix[,6:20], 1, mean) - apply(count_matrix[,1:5], 1, mean)

sort(abs(count_matrix_mean_diff), decreasing = TRUE)[1:5]
sort((count_matrix_mean_diff))[1]
```




```{r E3-Q2.4}
## Add your code here
#length(unique(names(unlist(ensembl2kegg_pathway)[which((unlist(ensembl2kegg_pathway) == "04010") == TRUE)])))
pathview(gene.data = as.data.frame(count_matrix_mean_diff), out.suffix = "MAPK_signaling", limit = list(gene=2, cpd=1), pathway.id = "04010", gene.idtype = "ENSEMBL")
#data(golub)
#golDiff=apply(golub[,1:27],1,median)-apply(golub[,28:38],1,median) 
#pv.out=pathview(gene.data=golDiff, pathway.id="04062",species="hsa", limit=list(gene=2,cpd=1), out.suffix="golDiff", #kegg.native=T)

```



## Exercise 4. R script interpretation

```{r E4-Q1.1}
library(org.Hs.eg.db)
#class(org.Hs.egPATH)
#typeof(org.Hs.egPATH)
map_ke <- as.list(org.Hs.egPATH)

kegg_entrez_pathway = map_ke[!is.na(match(mappings_ensembl_entrez$entrezgene_id, names(map_ke)))]
#typeof(kegg_entrez_pathway)
#nrow(mappings_ensembl_entrez)
#class(mappings_ensembl_entrez[2, ])
#mappings_ensembl_entrez[2, ]$entrezgene_id
#typeof(mappings_ensembl_entrez[2, ]$entrezgene_id)
#typeof(kegg_entrez_pathway[as.character(mappings_ensembl_entrez[5, ]$entrezgene_id)])
#!is.na(names(temp))
#a = kegg_entrez_pathway[as.character(mappings_ensembl_entrez[4045, ]$entrezgene_id)]
#!is.na(names(a))
#mappings_ensembl_entrez[which((is.na(mappings_ensembl_entrez$entrezgene_id)) == TRUE),]

ensembl2kegg_pathway = list()
for (i in 1:nrow(mappings_ensembl_entrez)) {
  entrez_id_i = mappings_ensembl_entrez[i, ]$entrezgene_id
  kegg_entrez_id_i = kegg_entrez_pathway[as.character(entrez_id_i)]
  names_kegg_entrez_id_i = names(kegg_entrez_id_i)
  
  if (!is.na(names_kegg_entrez_id_i)){
    ensembl2kegg_pathway[mappings_ensembl_entrez[i, ]$ensembl_gene_id] = kegg_entrez_id_i
  } 
}
```

```{r E4-Q.1.1 refactored}
library(org.Hs.eg.db)

map_ke <- as.list(org.Hs.egPATH)
kegg_entrez_pathway = map_ke[!is.na(match(mappings_ensembl_entrez$entrezgene_id, names(map_ke)))]

ensembl2kegg_pathway = list()

for (i in 1:nrow(mappings_ensembl_entrez)) {
  entrez_id_i = mappings_ensembl_entrez[i, ]$entrezgene_id
  kegg_entrez_id_i = kegg_entrez_pathway[as.character(entrez_id_i)]
  
  names_kegg_entrez_id_i = names(kegg_entrez_id_i)

  if (!is.na(names_kegg_entrez_id_i)){
    ensembl2kegg_pathway[mappings_ensembl_entrez[i, ]$ensembl_gene_id] = kegg_entrez_id_i
  } 
}
```


## Exercise 5. Pathway enrichment analysis; many pathways

### Calculate n for all pathways

```{r E5-Q1.1}
## Add your code here
ns = table(unlist(ensembl2kegg_pathway))
ns
```



### Calculate k for all pathways

```{r E5-Q1.2}
## Add your code here
ks = rep(0, length(ns))
ks_res = table(unlist(ensembl2kegg_pathway_de))
names(ks) = names(ns)
ks[names(ks_res)] = ks_res
ks
```



### Calculate pathway enrichment P-values

```{r E5-Q1.3}
## Add your code here
p_vals = 1 - phyper(ks - 1, M, N-M, ns)
round(sort(p_vals)[[1]], 5)
```


### Significant pathways

```{r E5-Q1.5}
## Add your code here
sum(p_vals < 0.05)
```



```{r E5-Q1.7}
## Add your code here
enriched_p_vals = sort(p_vals[p_vals < 0.05])
enriched_p_vals_df = data.frame(pathway_identifiers = names(enriched_p_vals), p_value = enriched_p_vals)
enriched_p_vals_df
```



```{r E5-Q1.9}
## Add your code here
VEGF_id = "04370"
round(p_vals[VEGF_id], 3)
```

